Wallace Session 2024-06-09
================

Please find below the R code history from your *Wallace* v2.1.1 session.

You can reproduce your session results by running this R Markdown file
in RStudio.

Each code block is called a “chunk”, and you can run them either
one-by-one or all at once by choosing an option in the “Run” menu at the
top-right corner of the “Source” pane in RStudio.

For more detailed information see <http://rmarkdown.rstudio.com>).

### Package installation

Wallace uses the following R packages that must be installed and loaded
before starting.

```{r}
library(spocc)
library(spThin)
library(dismo)
library(sf)
library(ENMeval)
library(wallace)
```

The *Wallace* session code .Rmd file is composed of a chain of module
functions that are internal to *Wallace*. Each of these functions
corresponds to a single module that the user ran during the session. To
see the internal code for these module functions, click on the links in
the .Rmd file. Users are encouraged to write custom code in the .Rmd
directly to modify their analysis, and even modify the module function
code to further customize. To see the source code for any module
function, just type its name into the R console and press Return.

```{r}
# example:
# just type the function name and press Return to see its source code
# paste this code into a new script to edit it
occs_queryDb
```

Your analyses are below.

------------------------------------------------------------------------

## Analysis for *Vulpes vulpes* (Vv)

### Obtain Occurrence Data

You searched the gbif database for *Vulpes vulpes*, limited to 1000
records. You decided to remove occurrences without uncertainty
information? TRUE

```{r}
# Query selected database for occurrence records
queryDb_Vv <- occs_queryDb(
  spNames = "Vulpes vulpes", 
  occDb = "gbif", 
  occNum = 1000,
  RmUncertain = TRUE)
occs_Vv <- queryDb_Vv$Vulpes_vulpes$cleaned
```

### Obtain environmental data

Using Ecoclimate (<http://www.ecoclimate.org>) dataset at resolution of
0.5 degrees.

```{r}
# R code to get environmental data from Ecoclimate
envs_Vv <- envs_ecoClimate(
  bcAOGCM = "CCSM",
  bcScenario = "Present",
  ecoClimSel = c(1, 3, 4, 6, 15)) 
##Add envrionmental values to occurrences table
occs_xy_Vv <- occs_Vv[c('longitude', 'latitude')]
occs_vals_Vv <- as.data.frame(raster::extract(envs_Vv, occs_xy_Vv, cellnumbers = TRUE))
# Remove duplicated same cell values
occs_Vv <- occs_Vv[!duplicated(occs_vals_Vv[, 1]), ]
occs_vals_Vv <- occs_vals_Vv[!duplicated(occs_vals_Vv[, 1]), -1]
# remove occurrence records with NA environmental values
occs_Vv <- occs_Vv[!(rowSums(is.na(occs_vals_Vv)) >= 1), ]
# also remove variable value rows with NA environmental values
occs_vals_Vv <- na.omit(occs_vals_Vv)
# add columns for env variable values for each occurrence record
occs_Vv <- cbind(occs_Vv, occs_vals_Vv)
```

### Process Occurrence Data

Remove occurrences outside of user drawn polygon

```{r}
occs_Vv <- poccs_selectOccs(
  occs = occs_Vv,
  polySelXY = matrix(c(-141.322632, -122.689819, 62.232056, 143.442993, 165.591431, 121.294556, -66.439819, -126.205444, -141.322632, 39.368279, 67.609221, 70.140364, 57.704147, 40.446947, 12.554564, 21.289374, 18.312811, 39.368279), ncol = 2, byrow = FALSE),
  polySelID = 5038)
```

### Process environmental data

Sampling of 1000 background points and corresponding environmental data
using a “point buffers” method with a 20 degree buffer.

```{r}
# Generate background extent 
bgExt_Vv <- penvs_bgExtent(
  occs = occs_Vv,
  bgSel = "point buffers",
  bgBuf = 20)
# Mask environmental data to provided extent
bgMask_Vv <- penvs_bgMask(
  occs = occs_Vv,
  envs = envs_Vv,
  bgExt = bgExt_Vv)
# Sample background points from the provided area
bgSample_Vv <- penvs_bgSample(
  occs = occs_Vv,
  bgMask =  bgMask_Vv,
  bgPtsNum = 1000)
# Extract values of environmental layers for each background point
bgEnvsVals_Vv <- as.data.frame(raster::extract(bgMask_Vv,  bgSample_Vv))
##Add extracted values to background points table
bgEnvsVals_Vv <- cbind(scientific_name = paste0("bg_", "Vulpes vulpes"), bgSample_Vv,
                            occID = NA, year = NA, institution_code = NA, country = NA,
                            state_province = NA, locality = NA, elevation = NA,
                            record_type = NA, bgEnvsVals_Vv)
```

### Partition occurrence data

Partition occurrences and background points for model training and
validation using random k-fold, a non-spatial partition method.

```{r}
# R code to get partitioned data
groups_Vv <- part_partitionOccs(
  occs = occs_Vv ,
  bg =  bgSample_Vv, 
  method = "rand",
  kfolds = 4) 
```

### Build and Evaluate Niche Model

Generating a species distribution model using the maxnet algorithm as
implemented in ENMeval V2.0 (with clamping = FALSE). For tuning using L,
LQH feature classes and regularization multipliers in the 1, 4 range
increasing by 3. Not using any categorical predictor variables.

```{r}
# Run maxent model for the selected species
model_Vv <- model_maxent(
  occs = occs_Vv,
  bg = bgEnvsVals_Vv,
  user.grp = groups_Vv, 
  bgMsk = bgMask_Vv,
  rms = c(1, 4), 
  rmsStep =  3,
  fcs = c('L', 'LQH'),
  clampSel = FALSE,
  algMaxent = "maxnet",
  parallel = TRUE,
  numCores = 6)
```

### Visualize

Generate a map of the maxnet generated model with no threshold

```{r}
# Select current model and obtain raster prediction
m_Vv <- model_Vv@models[["fc.LQH_rm.4"]]
predSel_Vv <- predictMaxnet(m_Vv, bgMask_Vv,
                                          type = "cloglog", 
                                          clamp = FALSE)
#Get values of prediction
mapPredVals_Vv <- getRasterVals(predSel_Vv, "cloglog")
#Define colors and legend  
rasCols <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")
legendPal <- colorNumeric(rev(rasCols), mapPredVals_Vv, na.color = 'transparent')
rasPal <- colorNumeric(rasCols, mapPredVals_Vv, na.color = 'transparent')
#Generate map
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) 
m  %>%
  leaflet::addLegend("bottomright", pal = legendPal,
            title = "Predicted Suitability<br>(Training)",
            values = mapPredVals_Vv, layerId = "train",
            labFormat = reverseLabel(2, reverse_order = TRUE)) %>% 
  #add occurrence data
  addCircleMarkers(data = occs_Vv, lat = ~latitude, lng = ~longitude,
                   radius = 5, color = 'red', fill = TRUE, fillColor = "red",
                   fillOpacity = 0.2, weight = 2, popup = ~pop) %>% 
  ##Add model prediction
  addRasterImage(predSel_Vv, colors = rasPal, opacity = 0.7,
                 group = 'vis', layerId = 'mapPred', method = "ngb") %>%
 ##add background polygons
  addPolygons(data = bgExt_Vv,fill = FALSE,
              weight = 4, color = "blue", group = 'proj')
```

```{r}
# First must generate the transfer area according to the drawn polygon in the GUI
xfer_draw_Vv <-xfer_draw(
  polyXfXY = matrix(c(-170.848389, -168.739014, 179.307861, 192.667236, -116.707764, -116.707764, -170.848389, 86.031103, -77.915669, -77.157163, 85.831591, 85.982137, 85.982137, 86.031103),ncol=2,byrow=FALSE),
  polyXfID = 29033,
  drawXfBuf = 0)
# Create object of transfer variables
xferAreaEnvs_Vv <- envs_Vv
# Generate a transfer of the model to the desired area
xfer_area_Vv <- xfer_area(
  evalOut = model_Vv,
  curModel = "fc.LQH_rm.4",
  envs = xferAreaEnvs_Vv , 
  outputType = "cloglog",
  alg = "maxnet",
  clamp = FALSE,
  xfExt = xfer_draw_Vv) 
# store the cropped transfer variables
xferExt_Vv <- xfer_area_Vv$xferExt
 #map result
 
 ###Make map of transfer
bb_Vv <-  bgExt_Vv@bbox
bbZoom <- polyZoom(bb_Vv[1, 1], bb_Vv[2, 1], bb_Vv[1, 2], 
                   bb_Vv[2, 2], fraction = 0.05)
mapXferVals_Vv <- getRasterVals(xfer_area_Vv$xferArea,"cloglog")
rasCols_Vv <- c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")
# if no threshold specified
legendPal <- colorNumeric(rev(rasCols_Vv), mapXferVals_Vv, na.color = 'transparent')
rasPal_Vv <- colorNumeric(rasCols_Vv, mapXferVals_Vv, na.color = 'transparent')
m <- leaflet() %>% addProviderTiles(providers$Esri.WorldTopoMap) 
m %>%
  fitBounds(bbZoom[1], bbZoom[2], bbZoom[3], bbZoom[4]) %>%
  leaflet::addLegend("bottomright", pal = legendPal,
            title = "Predicted Suitability<br>(Transferred)",
            values = mapXferVals_Vv, layerId = 'xfer',
            labFormat = reverseLabel(2, reverse_order = TRUE)) %>%
# map model prediction raster and transfer polygon
  clearMarkers() %>% clearShapes() %>% removeImage('xferRas') %>%
  addRasterImage(xfer_area_Vv$xferArea, colors = rasPal_Vv, opacity = 0.7,
                 layerId = 'xferRas', group = 'xfer', method = "ngb") %>%
 ##add transfer polygon (user drawn area)
  addPolygons(data = xfer_draw_Vv, fill = FALSE,
              weight = 4, color = "blue", group = 'xfer')
```
